---
title: "ETC4500/5450 Advanced R Programming"
execute:
  cache: false
---

R is widely used as a tool for data analysis and one of the most popular programming languages. This unit delves into R from the programming aspect instead of using it as a data analysis tool. You will learn a variety of programming paradigms and delve into the foundations of R as a programming language.

## Learning outcomes

1. be familiar with the foundations of R programming
2. understand a variety of programming paradigms, including functional programming and object-oriented programming
3. utilise R’s functional and object-oriented programming systems
4. integrate the concept of metaprogramming to evaluate and construct new functions

## Teachers

* [**Rob J Hyndman**](https://robjhyndman.com). Chief Examiner. Email: [Rob.Hyndman@monash.edu](mailto:Rob.Hyndman@monash.edu)
* [**Mitchell O'Hara-Wild**](https://mitchelloharawild.com). Email: [Mitch.OHara-Wild@monash.edu](mailto:Mitch.OHara-Wild@monash.edu)
* [**Hadley Wickham**](https://hadley.nz), Week 2
* [**Tomasz Wozniak**](https://github.com/donotdespair), Week 11

## References

* [*Advanced R*, 2e](https://adv-r.hadley.nz/) by Hadley Wickham.
* [*Advanced R Solutions*](https://advanced-r-solutions.rbind.io/) by Malte Grosser, Henning Bumann & Hadley Wickham.
* [*The {targets} R package user manual*](https://books.ropensci.org/targets/) by Will Landau.
* [*R packages*, 2e](http://r-pkgs.org/) by Hadley Wickham & Jenny Bryan.

```{r}
#| label: load_packages
#| include: false
#| message: false
#| warning: false
#| echo: false
#| cache: false
library(tidyverse)
options(knitr.kable.NA = '')
source(here::here("course_info.R"))
week <- as.integer(1 + ((Sys.Date() - as_date(start_semester))/7))
```

## Recordings

* [Recordings](https://learning.monash.edu/mod/lti/view.php?id=3661834)

## Weekly schedule

* 3 hour workshop, Thursdays 6pm-9pm each week, in Woodside G16.

```{r}
#| label: schedule2
#| message: false
#| warning: false
#| echo: false
#| output: asis
schedule |>
  transmute(
    Date = format(Date + 3, "%d %b"),
    Show = !is.na(Week) & Week <= 12,
    Topic = if_else(!Show, Topic, glue::glue("[{Topic}](./week{Week}/index.html)")),
    Assessments = if_else(is.na(Assignment), Assignment, glue::glue("[{Assignment}]({File})"))
  ) |>
  select(-Show) |>
  knitr::kable(format = "markdown")
```

## Assessments

* [Assignment 1](assignments/A1.qmd): 10%
* [Assignment 2](assignments/A2.qmd): 25%
* [Assignment 3](assignments/A3.qmd): 25%
* [Game package 1](assignments/Game_package1.qmd): 10%
* [Game package 2](assignments/Game_package2.qmd): 30%
