
---
title: "OTexts.com views by country"
date: "2026-03-26"
format: html
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

```{r}
#| label: load-packages
library(tidyverse)
```

```{r}
#| label: load-data
views <- read_csv(
  here::here("week4/views_by_country.csv"),
  comment = "#"
)
```

## Introduction

This report summarises website traffic to OTexts.com by country, based on Google Analytics 4 (GA4) data spanning January 2021 to January 2026. The dataset covers `{r} nrow(views)` countries and includes metrics such as views, active users, views per active user, average engagement time, and event count.


## Overview

Across all countries, OTexts.com recorded a total of **`{r} scales::comma(sum(views$Views))`** views from **`{r} scales::comma(sum(views$"Active users"))`** active users over the period.

```{r}
#| label: overview-stats
tibble(
  Metric = c("Countries", "Total views", "Total active users"),
  Value = c(
    scales::comma(nrow(views)),
    scales::comma(sum(views$Views)),
    scales::comma(sum(views$`Active users`))
  )
)
```

## Top 20 countries by views

The table below shows the 20 countries with the highest number of views, along with their active user counts and views per active user.

```{r}
#| label: top20-table
views |>
  arrange(desc(Views)) |>
  head(20) |>
  select(Country, Views, `Active users`, `Views per active user`) |>
  mutate(
    Views = scales::comma(Views),
    `Active users` = scales::comma(`Active users`),
    `Views per active user` = round(`Views per active user`, 1)
  )
```

## Bar chart: top 20 countries by views

The chart below shows the top 20 countries by total views, with bars coloured by continent.

```{r}
#| label: top20-bar-chart
#| fig-alt: "Horizontal bar chart of the top 20 countries by views on OTexts.com, with bars coloured by continent."
continent_map <- c(
  "United States" = "Americas",
  "India" = "Asia",
  "Australia" = "Oceania",
  "United Kingdom" = "Europe",
  "China" = "Asia",
  "South Korea" = "Asia",
  "Canada" = "Americas",
  "Germany" = "Europe",
  "Brazil" = "Americas",
  "Italy" = "Europe",
  "Malaysia" = "Asia",
  "Netherlands" = "Europe",
  "Vietnam" = "Asia",
  "Spain" = "Europe",
  "Japan" = "Asia",
  "Mexico" = "Americas",
  "Indonesia" = "Asia",
  "Singapore" = "Asia",
  "France" = "Europe",
  "Hong Kong" = "Asia"
)

top20 <- views |>
  arrange(desc(Views)) |>
  head(20) |>
  mutate(
    Continent = continent_map[Country],
    Country = fct_reorder(Country, Views)
  )

ggplot(top20, aes(x = Views, y = Country, fill = Continent)) +
  geom_col() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Top 20 countries by views",
    subtitle = "OTexts.com — Jan 2021 to Jan 2026",
    x = "Views",
    y = NULL,
    fill = "Continent"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

The United States dominates with approximately `{r} scales::comma(top20$Views[top20$Country == "United States"])` views — around three times that of the next country, India. Asian and European countries are well represented across the rest of the top 20, while Australia stands out as a high-traffic country relative to its population.

## Engagement

Views per active user provides a measure of how deeply visitors engage with the site. Among large countries (at least 1,000 views), Australia has notably high engagement.

```{r}
#| label: high-engagement
views |>
  filter(Views >= 1000) |>
  arrange(desc(`Views per active user`)) |>
  select(Country, Views, `Active users`, `Views per active user`) |>
  mutate(
    Views = scales::comma(Views),
    `Active users` = scales::comma(`Active users`),
    `Views per active user` = round(`Views per active user`, 1)
  ) |>
  head(10)
```
