---
title: "Week 12: Rewriting R code in C++"
---

```{r}
week <- 12
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```


## What you will learn this week

* **C++** for applications in **R** with [Tomasz Woźniak](https://github.com/donotdespair)
* The first steps with **Rcpp**
* Some stats with **RcppArmadillo**
* Hands-on coding with four exercises
* Create an **R** package with compiled code in ten steps

## Other references used during the session

* **Armadillo** library [documentation](https://arma.sourceforge.net/docs.html)
* Tsuda, M., [*Rcpp for everyone*](https://teuder.github.io/rcpp4everyone_en/)

## Scripts used during the session

* Simple list creation [`nicetry.cpp`](https://github.com/numbats/arp/blob/main/week12/nicetry.cpp)
* Simple linear regression example [`nicelr.cpp`](https://github.com/numbats/arp/blob/main/week12/nicelr.cpp)
* Simple loop application [`nicelist.cpp`](https://github.com/numbats/arp/blob/main/week12/nicelist.cpp)
* Sampling from inverted gamma 2 distribution [`nicerig2.cpp`](https://github.com/numbats/arp/blob/main/week12/nicerig2.cpp)
* Step-by-step package creation [`nicepackage.R`](https://github.com/numbats/arp/blob/main/week12/nicepackage.R)

```{r}
#| output: asis
show_slides(week)
show_assignments(week)
```
