---
title: ETC4500/ETC5450 Advanced&nbsp;R&nbsp;programming
author: "Week 1: Foundations of R programming"
format:
  presentation-beamer:
    fontsize: "14pt,t"
    section-titles: false
    knitr:
      opts_chunk:
        dev: "cairo_pdf"
    fig-width: 7.5
    fig-height: 3.5
    include-in-header: ../header.tex
    keep-tex: false
---

```{r setup, include=FALSE}
source(here::here("setup.R"))
source(here::here("course_info.R"))
```

# Unit admin

## Introduction

### Expectations

* You know R and RStudio
* You have a basic understanding of programming (for loops, if statements, functions)
* You can use Git and GitHub (https://happygitwithr.com)

### Unit resources

* Everything on **https://arp.numbat.space**
* Assignments submitted on Github Classroom
* Discussion on Ed

## Introduction

### Assessment

* Three group assignments building or fixing packages: 30% each
* Individual discussion participation: 10%

### GitHub

* Use your monash edu address.
* Apply to GitHub Education as a student (<https://github.com/education/students>).
* Gives you free access to private repos and GitHub Copilot.
* Add GitHub Copilot to RStudio settings, or sign into GitHub in Positron.

# The R language and tools

## R history

* S (1976, Chambers, Becker and Wilks; Bell Labs, USA)
* S-PLUS (1988, Doug Martin; Uni of Washington, USA)
* R (1993, Ihaka and Gentleman; Uni of Auckland, NZ)

\pause\vspace*{0.4cm}

### R influenced by

* Lisp (functional programming, environments, dynamic typing)
* Scheme (functional programming, lexical scoping)
* S and S-PLUS (syntax)

## Why R?

* Free, open source, and on every major platform.
* A diverse and welcoming community
* A massive set of packages, often cutting-edge.
* Powerful communication tools (Shiny, Rmarkdown, \rlap{quarto)}
* RStudio and Positron IDEs
* Deep-seated language support for data analysis.
* A strong foundation of functional programming.
* Posit
* Easy connection to high-performance programming languages like C, Fortran, and C++.

## R challenges

* R users are not usually programmers. Most R code by ordinary users is not very elegant, fast, or easy to understand.
* R users more focused on results than good software practices.
* R packages are inconsistent in design.
* R can be slow.

## RStudio and Positron

* Integrated development environments (IDEs) for R.
* Provide a user-friendly interface for writing and running R code, managing files, and visualizing data.
* RStudio is the most popular IDE for R, but Positron is a newer alternative that offers some additional features and is designed to also work with Python.
* GitHub integration is available in both IDEs.

## CRAN, Bioconductor, and GitHub

* CRAN: Comprehensive R Archive Network, the official repository for R packages.
* Bioconductor: a repository for R packages focused on bioinformatics.
* GitHub: a platform for hosting and sharing code, including R packages.

### Package installation

`pak::pak()` is a modern package installer that resolves dependencies and installs packages from CRAN, Bioconductor, and GitHub.


# Scalars and vectors

## Scalars

* **Logicals**: `TRUE` or `FALSE`, or abbreviated (`T` or `F`).

* **Doubles**: decimal (`0.1234`), scientific (`1.23e4`), or hexadecimal (`0xcafe`). Special values: `Inf`, `-Inf`, and `NaN` (not a number).

* **Integers**: `1234L`, `1e4L`, or `0xcafeL`. Can not contain fractional values.

* **Strings**: `"hi"` or `'bye'`. Special characters are escaped with `\`.

## Making longer vectors with `c()` {#atomic-constructing}

Use `c()` to create vectors.

```{r}
lgl_var <- c(TRUE, FALSE)
int_var <- c(1L, 6L, 10L)
dbl_var <- c(1, 2.5, 4.5)
chr_var <- c("these are", "some strings")
```
When the inputs are atomic vectors,\newline `c()` always flattens.

```{r}
c(c(1, 2), c(3, 4))
```

\placefig{9.8}{2.5}{width=6cm}{../diagrams/vectors/atomic.png}

## Atomic vectors
\vspace*{-0.2cm}

* Four primary types of atomic vectors: `logical`, `double`, `integer`,  and `character` (which contains strings).
* Two rare types: `complex`, `raw`.
* Collectively `integer` and `double` vectors are known as `numeric` vectors
* `NULL` is like a zero length vector
* Scalars are just vectors of length 1
* Every vector can also have __attributes__: a named list of arbitrary metadata.
* The __dimension__ attribute turns vectors into matrices and arrays.
* The __class__ attribute powers the S3 object system.


## Types and length
\fontsize{13}{13}\sf

You can determine the type of a vector with `typeof()` and its length with `length()`.

\fontsize{10}{10}\sf


```{r}
typeof(lgl_var)
typeof(int_var)
typeof(dbl_var)
typeof(chr_var)
```

## Missing values

Most computations involving a missing value will return another missing value.

\fontsize{10}{10}\sf

```{r}
NA > 5
10 * NA
!NA
```

## Missing values

Exceptions:

\fontsize{10}{10}\sf

```{r}
NA ^ 0
NA | TRUE
NA & FALSE
```

## Missing values

Use `is.na()` to check for missingness

\fontsize{10}{10}\sf

```{r}
x <- c(NA, 5, NA, 10)
x == NA
is.na(x)
```

\fontsize{13}{14}\sf

There are actually four missing values:  `NA` (logical), `NA_integer_` (integer), `NA_real_` (double), and `NA_character_` (character).

## Coercion
\fontsize{13}{14}\sf\vspace*{-0.25cm}

* For atomic vectors, all elements must be the same type.
* When you combine different types they are __coerced__ in a fixed order: logical → integer → double → character.

\fontsize{10}{4}\sf

```{r}
str(c("a", 1))
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)
sum(x)
as.integer(c("1", "1.5", "a"))
```

## Exercises

1.  Predict the output of the following:

    ```{r, eval=FALSE}
    c(1, FALSE)
    c("a", 1)
    c(TRUE, 1L)
    ```

2. Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?

3. Why is the default missing value, `NA`, a logical vector? What's special
   about logical vectors? (Hint: think about `c(FALSE, NA_character_)`.)

## Getting and setting attributes

* You can think of attributes as name-value pairs that attach metadata to an object.
* Individual attributes can be retrieved and modified with `attr()`, or retrieved en masse with `attributes()`, and set en masse with `structure()`.

\fontsize{10}{10}\sf

```{r}
a <- 1:3
attr(a, "x") <- "abcdef"
a
```

## Getting and setting attributes
\fontsize{10}{10}\sf

```{r}
attr(a, "y") <- 4:6
str(attributes(a))

# Or equivalently
a <- structure(
  1:3,
  x = "abcdef",
  y = 4:6
)
str(attributes(a))
```

\placefig{10}{1}{width=3.5cm}{../diagrams/vectors/attr.png}

## Names {#attr-names}

* Names are a type of attribute.
* You can name a vector in three ways:

\fontsize{10}{10}\sf

```{r}
# When creating it:
x <- c(a = 1, b = 2, c = 3)

# By assigning a character vector to names()
x <- 1:3
names(x) <- c("a", "b", "c")

# Inline, with setNames():
x <- setNames(1:3, c("a", "b", "c"))
```

```{r}
x
```

## Names

* Avoid using `attr(x, "names")` as it requires more typing and is less readable than `names(x)`.
* You can remove names from a vector by using\newline `x <- unname(x)` or `names(x) <- NULL`.


## Dimensions {#attr-dims}

* Adding a `dim` attribute to a vector allows it to behave like a 2-dimensional __matrix__ or a multi-dimensional __array__.
* You can create matrices and arrays with `matrix()` and `array()`, or by using the assignment form of `dim()`:

```{r}
# Two scalar arguments specify row and column sizes
x <- matrix(1:6, nrow = 2, ncol = 3)
x
```

## Dimensions

```{r}
# One vector argument to describe all dimensions
y <- array(1:12, c(2, 3, 2))
y
```

## Dimensions

```{r}
# You can also modify an object in place by setting dim()
z <- 1:6
dim(z) <- c(3, 2)
z
```

## Exercises

4. What does `dim()` return when applied to a 1-dimensional vector?
5. When might you use `NROW()` or `NCOL()`?
6. How would you describe the following three objects? What makes them different from `1:5`?

   ```{r}
   x1 <- array(1:5, c(1, 1, 5))
   x2 <- array(1:5, c(1, 5, 1))
   x3 <- array(1:5, c(5, 1, 1))
   ```

## S3 atomic vectors

* `class` is a vector attribute.
* It turns object into __S3 object__.
* Four important S3 vectors:

  * **factor** vectors.
  * **Date** vectors with day resolution.
  * **POSIXct** vectors for date-times.
  * **difftime** vectors for durations.

\placefig{10}{2}{width=6cm}{../diagrams/vectors/summary-tree-s3-1.png}

## Factors

* A vector that can contain only predefined values.
* Used to store categorical data.
* Built on top of an integer vector with two attributes: a `class`, "factor", and `levels`, which defines the set of allowed values.

```{r}
x <- factor(c("a", "b", "b", "a"))
x
```

## Factors

```{r}
typeof(x)
attributes(x)
```

\placefig{10}{2}{width=5cm}{../diagrams/vectors/factor.png}

## Factors

```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

## Factors

* Be careful: some functions convert factors to integers!
* ggplot preserves ordering of levels in graphs -- useful to reorder panels or legends.
* Ordered factors are useful when the order of levels is meaningful.

```{r}
grade <- ordered(c("b", "b", "a", "c"), levels = c("c", "b", "a"))
grade
```

## Dates

* Date vectors are built on top of double vectors.
* Class "Date" with no other attributes:

```{r}
today <- Sys.Date()

typeof(today)
attributes(today)
```

## Dates

The value of the double (which can be seen by stripping the class), represents the number of days since 1970-01-01 (the "Unix Epoch").

```{r}
date <- as.Date("1970-02-01")
unclass(date)
```

## Date-times

* Base R provides two ways of storing date-time information, POSIXct, and POSIXlt.
* "POSIX" is short for Portable Operating System Interface
* "ct" stands for calendar time; "lt" for local time
* POSIXct vectors are built on top of double vectors, where the value represents the number of seconds since 1970-01-01.

```{r}
now_ct <- as.POSIXct("2018-08-01 22:00", tz = "UTC")
now_ct

typeof(now_ct)
attributes(now_ct)
```

## Date-times

The `tzone` attribute controls only how the date-time is formatted; it does not control the instant of time represented by the vector. Note that the time is not printed if it is midnight.

\fontsize{10}{10}\sf

```{r}
structure(now_ct, tzone = "Asia/Tokyo")
structure(now_ct, tzone = "America/New_York")
structure(now_ct, tzone = "Australia/Lord_Howe")
```


## Exercises

7. What sort of object does `table()` return? What is its type? What
    attributes does it have? How does the dimensionality change as you
    tabulate more variables?

8. What happens to a factor when you modify its levels?

   ```{r, results = FALSE}
   f1 <- factor(letters)
   levels(f1) <- rev(levels(f1))
   ```

9. What does this code do? How do `f2` and `f3` differ from `f1`?

   ```{r, results = FALSE}
   f2 <- rev(factor(letters))
   f3 <- factor(letters, levels = rev(letters))
   ```

# Lists and data frames
## Lists

* More complex than atomic vectors
* Elements are *references* to objects of any type

\fontsize{10}{10}\sf

```{r}
l1 <- list(
  1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9)
)
typeof(l1)
str(l1)
```

\placefig{10}{3.5}{width=5cm}{../diagrams/vectors/list.png}

## Lists

* Lists can be recursive:  a list can contain other lists.

```{r}
l3 <- list(list(list(1)))
str(l3)
```

\placefig{10}{4}{width=5cm}{../diagrams/vectors/list-recursive.png}

## Lists

* `c()` will combine several lists into one.

\fontsize{9}{9}\sf

```{r}
l4 <- list(list(1, 2), c(3, 4))
l5 <- c(list(1, 2), c(3, 4))
str(l4)
str(l5)
```

\placefig{10}{3}{width=5cm}{../diagrams/vectors/list-c.png}

## Testing and coercion {#list-types}

\fontsize{9}{9}\sf

```{r}
list(1:3)
as.list(1:3)
```

\fontsize{13}{14}\sf

* You can turn a list into an atomic vector with `unlist()`.

## Data frames and tibbles {#tibble}
\vspace*{-0.2cm}
\placefig{12.2}{0.35}{width=3cm}{../diagrams/vectors/summary-tree-s3-2.png}

* Most important S3 vectors built on lists:\newline data frames and tibbles.

\fontsize{10}{8}\sf

```{r}
df1 <- data.frame(x = 1:3, y = letters[1:3])
typeof(df1)
attributes(df1)
```


\placefig{5.5}{5.5}{width=2.5cm}{../diagrams/vectors/data-frame-2.png}


## Data frames and tibbles

* A data frame has a constraint: the length of each of its vectors must be the same.
* A data frame has `rownames()` and `colnames()`. The `names()`
  of a data frame are the column names.
* A data frame has `nrow()` rows and `ncol()` columns. The `length()` of a
  data frame gives the number of columns.

## Tibbles
\fontsize{13}{14}\sf

* Modern reimagining of the data frame.
* tibbles are "lazy and surly": they do less and complain more.

\fontsize{10}{8}\sf

```{r}
library(tibble)
df2 <- tibble(x = 1:3, y = letters[1:3])
typeof(df2)
attributes(df2)
```


## Creating data frames and tibbles

```{r}
names(data.frame(`1` = 1))

names(tibble(`1` = 1))
```

## Creating data frames and tibbles

\fontsize{10}{10}\sf

```{r, error = TRUE}
data.frame(x = 1:4, y = 1:2)
tibble(x = 1:4, y = 1:2)
```

## Creating data frames and tibbles
\fontsize{10}{10}\sf

```{r}
tibble(
  x = 1:3,
  y = x * 2,
  z = 5
)
```


## Row names {#rownames}

Data frames allow you to label each row with a name, a character vector containing only unique values:

\fontsize{10}{10}\sf

```{r}
df3 <- data.frame(
  age = c(35, 27, 18),
  hair = c("blond", "brown", "black"),
  row.names = c("Bob", "Susan", "Sam")
)
df3
```

## Row names

* tibbles do not support row names
* convert row names into a regular column with either `rownames_to_column()`, or the `rownames` argument:

\fontsize{10}{10}\sf

```{r}
as_tibble(df3, rownames = "name")
```

## Printing

\fontsize{10}{9.5}\sf

```{r}
dplyr::starwars
```

## Printing

* Tibbles only show first 10 rows and all columns that fit on
  screen. Additional columns shown at bottom.
* Each column labelled with its type, abbreviated to 3--4 letters.
* Wide columns truncated.

## List columns

\fontsize{10}{10}\sf

```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)
df
```

\placefig{10}{1.5}{width=5cm}{../diagrams/vectors/data-frame-list.png}


```{r}
tibble(
  x = 1:3,
  y = list(1:2, 1:3, 1:4)
)
```

## Matrix and data frame columns

```{r}
dfm <- tibble(
  x = 1:3 * 10,
  y = matrix(1:9, nrow = 3),
  z = data.frame(a = 3:1, b = letters[1:3])
)
str(dfm)
```

\placefig{10}{1.25}{width=5cm}{../diagrams/vectors/data-frame-matrix.png}

## Exercises

10.  What happens if you attempt to set rownames that are not unique?
11.  If `df` is a data frame, what can you say about `t(df)`, and `t(t(df))`?
    Perform some experiments, making sure to try different column types.
12.  What does `as.matrix()` do when applied to a data frame with
    columns of different types? How does it differ from `data.matrix()`?

## `NULL`

```{r, error = TRUE}
length(NULL)
```

You can test for `NULL`s with `is.null()`:

```{r}
x <- NULL
x == NULL
is.null(x)
```

