---
title: ETC4500/ETC5450 Advanced&nbsp;R&nbsp;programming
author: "Week 2: Foundations of R programming"
format:
  presentation-beamer:
    fontsize: "14pt,t"
    section-titles: false
    toc: true
    knitr:
      opts_chunk:
        dev: "cairo_pdf"
    fig-width: 7.5
    fig-height: 3.5
    include-in-header: ../header.tex
    keep-tex: false
---

```{r setup, include=FALSE}
source(here::here("setup.R"))
source(here::here("course_info.R"))
```


# Subsetting

## Exercises

1. What is the result of subsetting a vector with positive integers, negative integers, a logical vector, or a character vector?
2. What's the difference between `[`, `[[`, and `$` when applied to a list?
3. When should you use `drop = FALSE`?

## Exercises
\fontsize{13}{14}\sf

4. Fix each of the following common data frame subsetting errors:

   ```{r}
   #| eval: false
   mtcars[mtcars$cyl = 4, ]
   mtcars[-1:4, ]
   mtcars[mtcars$cyl <= 5]
   mtcars[mtcars$cyl == 4 | 6, ]
   ```

5. Extract the residual degrees of freedom from `mod`

   ```{r}
   #| eval: false
   mod <- lm(mpg ~ wt, data = mtcars)
   ```

6. Extract the R squared from the model summary (`summary(mod)`)

## Exercises

7. How would you randomly permute the columns of a data frame?

8. How would you select a random sample of m rows from a data frame? What if the sample had to be contiguous (i.e., with an initial row, a final row, and every row in between)?

9. How could you put the columns in a data frame in alphabetical order?

# Functions

## Pre-class Exercise

Write a function `linear`, that takes arguments `x`, `a` and `b`, and returns `a + b * x`. Give default values `a=1` and `b=0`, and make sure your function has sensible error messages when the inputs are not as expected. Your function should work when `x` is a numerical scalar, vector, or matrix.

## Function fundamentals

* Almost all functions can be broken down into three components: arguments, body, and environment.

  * The `formals()`, the list of arguments that control how you call the function.
  * The `body()`, the code inside the function.
  * The `environment()`, the data structure that determines how the function finds the values associated with the names.

* Functions are objects and have attributes.

## Function components
\fontsize{10}{5}\sf

```{r}
linear <- function(x, a = 0, b = 1) {
  stopifnot(is.numeric(x), is.numeric(a), is.numeric(b))
  stopifnot(length(a) == 1, length(b) == 1)
  a + b * x
}
```

```{r}
formals(linear)
```


## Function components
\fontsize{10}{5}\sf

```{r}
linear <- function(x, a = 0, b = 1) {
  stopifnot(is.numeric(x), is.numeric(a), is.numeric(b))
  stopifnot(length(a) == 1, length(b) == 1)
  a + b * x
}
```

```{r}
body(linear)
```

## Function components
\fontsize{10}{5}\sf

```{r}
linear <- function(x, a = 0, b = 1) {
  stopifnot(is.numeric(x), is.numeric(a), is.numeric(b))
  stopifnot(length(a) == 1, length(b) == 1)
  a + b * x
}
```

```{r}
environment(linear)
```

## Invoking a function
\fontsize{10}{10}\sf

```{r}
mean(1:10, na.rm = TRUE)
mean(, TRUE, x = 1:10)
args <- list(1:10, na.rm = TRUE)
do.call(mean, args)
```

## Function composition {#function-composition}
\fontsize{13}{15}\sf

```{r}
square <- function(x) { x^2 }
deviation <- function(x) { x - mean(x) }
x <- runif(100)
```

:::: {.columns}
::: {.column width="50%"}

Nesting:

```{r}
sqrt(mean(square(deviation(x))))
```

Intermediate variables:

```{r}
out <- deviation(x)
out <- square(out)
out <- mean(out)
out <- sqrt(out)
out
```
:::

::: {.column width="50%"}

Pipe:

```{r}
x |>
  deviation() |>
  square() |>
  mean() |>
  sqrt()
```

:::
::::

## Lexical scoping {#lexical-scoping}

Names defined inside a function mask names defined outside a function.

\fontsize{10}{10}\sf

```{r}
x <- 10
y <- 20
g02 <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
g02()
```

## Lexical scoping

Names defined inside a function mask names defined outside a function.

\fontsize{10}{10}\sf

```{r}
x <- 2
g03 <- function() {
  y <- 1
  c(x, y)
}
g03()

# And this doesn't change the previous value of y
y
```

## Lexical scoping

Names defined inside a function mask names defined outside a function.

\fontsize{10}{10}\sf

```{r}
x <- 1
g04 <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}
g04()
```

## Functions versus variables
\fontsize{10}{10}\sf

```{r}
g07 <- function(x) { x + 1 }
g08 <- function() {
  g07 <- function(x) { x + 100 }
  g07(10)
}
g08()
```

```{r}
g09 <- function(x) { x + 100 }
g10 <- function() {
  g09 <- 10
  g09(g09)
}
g10()
```


## A fresh start {#fresh-start}

What happens to values between invocations of a function?

```{r}
#| echo: false
rm(a)
```

\fontsize{10}{9}\sf


```{r}
g11 <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  a
}

g11()
g11()
```

## Dynamic lookup
\fontsize{10}{10}\sf

```{r}
g12 <- function() { x + 1 }
x <- 15
g12()

x <- 20
g12()
```

```{r}
codetools::findGlobals(g12)
```

\pause\alert{It is good practice to pass all the inputs to a function as arguments.}

## Lazy evaluation {#lazy-evaluation}

This code doesn't generate an error because `x` is never used:

```{r}
h01 <- function(x) {
  10
}
h01(stop("This is an error!"))
```

## Promises
\fontsize{13}{14}\sf

Lazy evaluation is powered by a data structure called a __promise__.

A promise has three components:

*   An expression, like `x + y`, which gives rise to the delayed computation.

*   An environment where the expression should be evaluated

*   A value, which is computed and cached the first time a promise is
    accessed when the expression is evaluated in the specified environment.

## Promises
\fontsize{10}{10}\sf

```{r}
y <- 10
h02 <- function(x) {
  y <- 100
  x + 1
}
h02(y)
```

## Promises
\fontsize{10}{10}\sf

```{r}
#| message: true
double <- function(x) {
  message("Calculating...")
  x * 2
}
h03 <- function(x) {
  c(x, x)
}
h03(double(20))
```

\pause\fontsize{13}{15}\sf

Promises are like a quantum state: any attempt to inspect them with R code will force an immediate evaluation, making the promise disappear.

## Default arguments

Thanks to lazy evaluation, default values can be defined in terms of other arguments, or even in terms of variables defined later in the function:

\fontsize{10}{10}\sf

```{r}
h04 <- function(x = 1, y = x * 2, z = a + b) {
  a <- 10
  b <- 100
  c(x, y, z)
}
h04()
```

\pause\alert{Not recommended!}

## `...`  (dot-dot-dot) {#fun-dot-dot-dot}

Allows for any number of additional arguments.

You can use `...` to pass additional arguments to another function.
\fontsize{10}{10}\sf

```{r}
i01 <- function(y, z) {
  list(y = y, z = z)
}
i02 <- function(x, ...) {
  i01(...)
}
str(i02(x = 1, y = 2, z = 3))
```

## `...`  (dot-dot-dot)

`list(...)` evaluates the arguments and stores them in a list:\fontsize{10}{10}\sf


```{r}
i04 <- function(...) {
  list(...)
}
str(i04(a = 1, b = 2))
```

## `...`  (dot-dot-dot)
\fontsize{13}{14}\sf

*   Allows you to pass arguments to a function called within your function, without having to list them all explicitly.

\pause\alert{Two downsides:}

*   When you use it to pass arguments to another function, you have to
    carefully explain to the user where those arguments go.
*   A misspelled argument will not raise an error. This makes it easy for
    typos to go unnoticed:

\fontsize{10}{5}\sf

```{r}
sum(1, 2, NA, na_rm = TRUE)
```

## Exercises

10. Explain the following results: \fontsize{10}{10}\sf

    ```{r}
    sum(1, 2, 3)
    mean(1, 2, 3)

    sum(1, 2, 3, na.omit = TRUE)
    mean(1, 2, 3, na.omit = TRUE)
    ```


## Exiting a function

Most functions exit in one of two ways:

* return a value, indicating success
* throw an error, indicating failure.

## Implicit versus explicit returns

Implicit return, where the last evaluated expression is the return value:

\fontsize{10}{5}\sf

```{r}
j01 <- function(x) {
  if (x < 10) {
    0
  } else {
    10
  }
}
j01(5)
j01(15)
```

## Implicit versus explicit returns

Explicit return, by calling `return()`:

\fontsize{10}{5}\sf

```{r}
j02 <- function(x) {
  if (x < 10) {
    return(0)
  } else {
    return(10)
  }
}
j02(5)
j02(15)
```

## Invisible values {#invisible}

Most functions return visibly: calling the function in an interactive context prints the result.

```{r}
j03 <- function() { 1 }
j03()
```

However, you can prevent automatic printing by applying `invisible()` to the last value:

```{r}
j04 <- function() { invisible(1) }
j04()
```

## Invisible values {#invisible2}

The most common function that returns invisibly is `<-`:

```{r}
a <- 2
(a <- 2)
```

This is what makes it possible to chain assignments:

```{r}
a <- b <- c <- d <- 2
```

In general, any function called primarily for a side effect (like `<-`, `print()`, or `plot()`) should return an invisible value (typically the value of the first argument).

## Errors

If a function cannot complete its assigned task, it should throw an error with `stop()`, which immediately terminates the execution of the function.

\fontsize{10}{10}\sf

```{r}
#| error: true
j05 <- function() {
  stop("I'm an error")
  return(10)
}
j05()
```



## Function forms

###
> To understand computations in R, two slogans are helpful:
>
> * Everything that exists is an object.
> * Everything that happens is a function call.
>
> --- John Chambers

## Function forms

* __prefix__: the function name comes before its arguments, like
  `foofy(a, b, c)`.
* __infix__: the function name comes in between its arguments, like
  `x + y`.
* __replacement__: functions that replace values by assignment, like
  `names(df) <- c("a", "b", "c")`.
* __special__: functions like `[[`, `if`, and `for`.

## Rewriting to prefix form {#prefix-transform}

Everything can be written in prefix form.

```{r}
#| eval: false
x + y
`+`(x, y)

names(df) <- c("x", "y", "z")
`names<-`(df, c("x", "y", "z"))

for(i in 1:10) print(i)
`for`(i, 1:10, print(i))
```

## Don't be evil!

```{r}
`(` <- function(e1) {
  if (is.numeric(e1) && runif(1) < 0.1) {
    e1 + 1
  } else {
    e1
  }
}
replicate(50, (1 + 2))
```

```{r}
#| echo: false
rm("(")
```

## Prefix form {#prefix-form}

You can specify arguments in three ways:

* By position, like `help(mean)`.
* By name, like `help(topic = mean)`.
* Using partial matching, like `help(top = mean)`.

## Exercises

11. Clarify the following list of odd function calls:

    ```{r}
    #| eval: false
    x <- sample(replace = TRUE, 20, x = c(1:10, NA))
    y <- runif(min = 0, max = 1, 20)
    cor(m = "k", y = y, u = "p", x = x)
    ```

## Infix functions

Functions with 2 arguments, and the function name comes between the arguments:

 `:`, `::`, `:::`, `$`, `@`, `^`, `*`, `/`, `+`, `-`, `>`, `>=`, `<`, `<=`, `==`, `!=`, `!`, `&`, `&&`, `|`, `||`, `~`, `<-`, and `<<-`.

```{r}
1 + 2
`+`(1, 2)
```

## Infix functions

You can also create your own infix functions that start and end with `%`.

```{r}
`%+%` <- function(a, b) paste0(a, b)
"new " %+% "string"
```

## Replacement functions {#replacement-functions}

* Replacement functions act like they modify their arguments in place, and have the special name `xxx<-`.
* They must have arguments named `x` and `value`, and must return the modified object.

```{r}
`second<-` <- function(x, value) {
  x[2] <- value
  x
}
x <- 1:10
second(x) <- 5L
x
```

## Replacement functions

```{r}
`modify<-` <- function(x, position, value) {
  x[position] <- value
  x
}
modify(x, 1) <- 10
x
```

When you write `modify(x, 1) <- 10`, behind the scenes R turns it into:

```{r}
#| eval: false
x <- `modify<-`(x, 1, 10)
```

# Environments

## Environment basics

Environments are data structures that power scoping.

An environment is similar to a named list, with four important exceptions:

* Every name must be unique.
* The names in an environment are not ordered.
* An environment has a parent.
* Environments are not copied when modified.

## Environment basics

```{r}
library(rlang)
e1 <- env(
  a = FALSE,
  b = "a",
  c = 2.3,
  d = 1:3,
)
```

\placefig{8}{1.9}{width=7cm}{../diagrams/environments/bindings.png}

### Special environments

* The **current environment** is the environment in which code is currently executing.
* The **global environment** is where all interactive computation takes place. Your "workspace".

## Parent environments

:::: {.columns}
::: {.column width="50%"}

* Every environment has a parent.
* If a name is not found in an environment, R looks in the parent environment.
* The ancestors of the global environment include every attached package.

:::
::: {.column width="50%"}
\fontsize{10}{10}\sf\vspace*{-0.3cm}

```{r}
env_parents(e1, last = empty_env())
```
:::
::::

## Super assignment

* Regular assignment (`<-`) creates a variable in the current environment.
* Super assignment (`<<-`) modifies a variable in a parent environment.
* If it can't find an existing variable, it creates one in the global environment.

## Package environments
\fontsize{13}{15}\sf

* Every package attached becomes one of the parents of the global environment (in order of attachment).\fontsize{10}{10}\sf

  ```{r}
  search()
  ```

\fontsize{13}{15}\sf

* Attaching a package changes the parent of the global environment.
* `Autoloads` uses delayed bindings to save memory by only loading package objects when needed.

## Function environments
\fontsize{13}{14}\sf

A function binds the current environment when it is created.

\fontsize{10}{10}\sf

```{r}
y <- 1
f <- function(x) {
  env_print(current_env())
  x + y
}
f(2)
```

## Caller environments
\fontsize{10}{10}\sf

```{r}
f <- function(x) {
  g(x = 2)
}
g <- function(x) {
  h(x = 3)
}
h <- function(x) {
  stop()
}
```

```{r}
#| eval: false
f(x = 1)
#> Error: in h(x = 3)
traceback()
#> 4: stop() at #3
#> 3: h(x = 3) at #3
#> 2: g(x = 2) at #3
#> 1: f(x = 1)
```

## Lazy evaluation

```{r}
#| eval: false
a <- function(x) b(x)
b <- function(x) c(x)
c <- function(x) x
a(f())
#> Error: in h(x = 3)
traceback()
#> 7: stop() at #3
#> 6: h(x = 3) at #3
#> 5: g(x = 2) at #3
#> 4: f() at #1
#> 3: c(x) at #1
#> 2: b(x) at #1
#> 1: a(f())
```


# Conditions

## Conditions

```r
message("This is what a message looks like")
#> This is what a message looks like

warning("This is what a warning looks like")
#> Warning: This is what a warning looks like

stop("This is what an error looks like")
#> Error in eval(expr, envir, enclos): This is what an error looks like
```

\pause

* Ignore messages with `suppressMessages()`.
* Ignore warnings with `suppressWarnings()`.
* Ignore errors with `try()`.

## try()

* Allows execution to continue even if an error occurs.
* Returns a special object that captures the error.

\fontsize{10}{10}\sf

:::: {.columns}
::: {.column width="50%"}

```{r}
#| error: true
f1 <- function(x) {
  log(x)
  10
}

f1("x")
```

:::

\pause

::: {.column width="50%"}

```{r}
#| error: true
f2 <- function(x) {
  try(log(x))
  10
}
f2("a")
```

:::
::::

## Handling conditions
\fontsize{13}{15}\sf

Allow you to specify what should happen when a condition occurs.

```{r}
#| eval: false
tryCatch(
  error = function(cnd) {
    # code to run when error is thrown
  },
  code_to_run_while_handlers_are_active
)
withCallingHandlers(
  warning = function(cnd) {
    # code to run when warning is signalled
  },
  message = function(cnd) {
    # code to run when message is signalled
  },
  code_to_run_while_handlers_are_active
)
```

## tryCatch()
\fontsize{10}{10}\sf

```{r}
f3 <- function(x) {
  tryCatch(
    error = function(cnd) NA,
    log(x)
  )
}

f3("x")
```


## withCallingHandlers()
\fontsize{10}{10}\sf

```{r}
f4 <- function(x) {
  withCallingHandlers(
    warning = function(cnd) cat("How did this happen?\n"),
    log(x)
  )
}

f4(-1)
```


## Activity

Write a function to take a single integer input and return:

* `fizz` if the number is divisible by 5
* `buzz` if the number is divisible by 7
* `fizzbuzz` if the number is divisible by both 5 and 7
* the number otherwise

Your function should contain a `stop()` if the input is not an integer.
