---
title: ETC4500/ETC5450 Advanced&nbsp;R&nbsp;programming
author: "Week 3: R package development"
format:
  beamer:
    pdf-engine: pdflatex
    aspectratio: 169
    fontsize: "14pt,t"
    section-titles: false
    knitr:
      opts_chunk:
        dev: "cairo_pdf"
    fig-width: 7.5
    fig-height: 3.5
    include-in-header: ../header.tex
    keep-tex: false
---

```{r}
#| label: setup
#| include: false
#| cache: false
source(here::here("setup.R"))
source(here::here("course_info.R"))
```

## Outline

\vspace*{0.4cm}
\tableofcontents

# Getting started

## System setup

```r
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```

\pause

### R build toolchain
\vspace*{-0.2cm}
* Windows: [`https://cran.r-project.org/bin/windows/Rtools/`](https://cran.r-project.org/bin/windows/Rtools/)
* macOS: `xcode-select --install`
* Linux: `sudo apt install r-base-dev`

\pause

### Verify
\vspace*{-0.2cm}
`devtools::dev_sitrep()`

## Package states

* **source**: the original files
* **bundled**: some processing, and compressed to a single `.tar.gz` file (e.g., to upload to CRAN)
* **binary**: what you usually download from CRAN
* **installed**: decompressed binary file stored in package library
* **in-memory**: loaded into R session using `library()`

## Package states

\fullheight{../diagrams/install-load.png}


## Package structure

\placefig{3}{1.4}{width = 20cm, height=7.6cm, trim = 30 420 740 200, clip = true}{../diagrams/package-development.pdf}

## Package name

* Only letters, numbers and periods.
* Must start with a letter.
* It cannot end with a period.
* No hyphens or underscores.
* Use the `available::available()` function to try ideas.

## Package code is different

* The `DESCRIPTION` file is the principal way to declare dependencies; we don’t do this via `library(somepackage)`.
* Be explicit about which functions are user-facing and which are internal helpers. By default, functions are not exported.

## Getting started

```{r}
#| eval: false
usethis::create_package()
```

## Workflow

\full{../diagrams/workflow.png}

## Workflow shortcuts

* `install()` : Ctrl-Shift-B
* `load_all()` : Ctrl-Shift-L
* `document()` : Ctrl-Shift-D
* `check()` : Ctrl-Shift-E
* `test()` : Ctrl-Shift-T

# Package metadata

## DESCRIPTION file
\fontsize{10}{10}\sf

```
Package: doubler
Title: This package doubles numbers
Version: 0.0.0.9000
Authors@R:
    person("Rob", "Hyndman", , "Rob.Hyndman@monash.edu", role = c("aut", "cre"))
Description: Whether the input is real, complex or character, this will double it.
License: GPL (>= 3)
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.3.1
Suggests:
    testthat (>= 3.0.0)
Config/testthat/edition: 3
```

## DESCRIPTION file

* **Title**: one line description. Plain text, title case, no more than 65 characters.
* **Description**: Several sentences, one paragraph. 80 characters per line, 4 space indentation. Don't include the package name in the Title or Description. Do not start with "This package does..."
* **Author**: Use `Authors@R` with `person()` for each author.
* **Version**. `Major.Minor.Patch.9000`. The `9000` is a placeholder for development versions.
* **License**: `GPL-3` or `MIT` are common.

## DESCRIPTION file

* **Depends**: packages that are attached with your package. (Not needed for most packages.)
* **Imports**: packages that are used in your package. (Refer to functions using `pkg::fun()`.)
* **Suggests**: packages that are used in your package, but not required. (E.g., in tests or examples.)
* **LazyData**: `true` prevents users using `data()`.

## DESCRIPTION file

Functions to help with the DESCRIPTION file:

* `use_github()` or `use_github_links()`: set the GitHub repository, URL and BugReports.
* `use_mit_license()`: set the license to MIT.
* `use_gpl3_license()`: set the license to GPL-3.
* `use_package()`: Add package to Imports or Suggests.
* `use_data()`: Add data to your package.
* `use_tidy_description()`: Clean up the DESCRIPTION file.

## NAMESPACE file

* Generated by `roxygen2`, so don't edit by hand.
* `export()`: export a function (including S3 and S4 generics).
* `S3method()`: export an S3 method.
* `importFrom()`: import selected object from another namespace (including S4 generics).
* `import()`: import all objects from another package’s namespace.
* `useDynLib()`: registers routines from a DLL (this is specific to packages with compiled code).

# Documentation

## Documenting the package

```r
use_package_doc()
```

## Documenting functions

* Add roxygen2 comments to your `.R` files
  * RStudio menu: `Code > Insert roxygen skeleton` (while cursor is within function)
  * Or use Github Copilot (in RStudio or VS-Code)
  * Or write them by hand
* Then use `document()` to generate the `Rd` files and the `NAMESPACE` file. (Or press `Ctrl+Shift+D` in RStudio.)
* Preview documentation with `?function`

## Documenting functions
\fontsize{9}{9}\sf

```r
#' Title
#'
#' Description
#' More description
#'
#' @param x Description of x
#' @inheritParams fun
#' @returns Description of return value
#' @examples
#' @importFrom pkg fun
#' @import pkg
#' @rdname fun
#' @aliases fun
#' @seealso fun
#' @references Some reference
#' @author Your name
#' @export
```

## Documenting data

* Put raw data in `data-raw/`
* Code to wrangle data and create objects in `data-raw/`
* `use_data(object)` to add rda to `data/`

```r
#' Title
#'
#' Description
#' More description
#'
#' @source Where did you get the data?
#' @format Class, dimensions, or other details
#' @keywords datasets
#' @examples
"object"
```

## Vignettes

* A long-form guide to your package, or an extended example.
* `usethis::use_vignette("my-vignette")`
  * Creates a `vignettes/` directory.
  * Adds the necessary dependencies to `DESCRIPTION`
  * Drafts a vignette, `vignettes/my-vignette.Rmd`.
  * Adds some patterns to `.gitignore`

## Vignettes YAML

```yaml
---
title: "Vignette Title"
author: Your name
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```

## Vignettes initial code chunks

````{verbatim}
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
````

````{verbatim}
```{r setup}
library(yourpackage)
```
````

* Any package used in a vignette must be included in Suggests if not already in Imports.


## README.Rmd

1. Describe the high-level purpose of the package.
2. A simple example illustrating package.
3. Installation instructions
4. An overview of the main components of the package.

* Like a short vignette
* Displayed on the Github repository and the front page of the pkgdown site.
* Create with `usethis::use_readme_rmd()`
* Build with `devtools::build_readme()`

## NEWS

* List changes in each release that users might care \rlap{about.}
* Use `usethis::use_news_md()` to create a `NEWS.md` file.

\fontsize{9}{9}\sf

```markdown
# foofy (development version)

* Better error message when grooving an invalid grobble (#206).

# foofy 1.0.0

## Major changes

* Can now work with all grooveable grobbles!

## Minor improvements and bug fixes

* Printing scrobbles no longer errors (@githubusername, #100).

* Wibbles are now 55% less jibbly (#200).
```

# Tests

## testthat v3

* `usethis::use_testthat(3)`
  * Create a `tests/testthat/` directory.
  * Add `testthat` to the Suggests field in DESCRIPTION and specify `testthat 3e` in the `Config/testthat/edition` field.
  * Create a file `tests/testthat.R` that runs all your tests when `check()` runs.
* Every exported function should have tests.
* `usethis::use_test("some_tests.R")` creates a test file for a function or group of functions.
* Each R file should match a test file.

## testthat v3

* Test files live in `tests/testthat/` and are named `test-*.R`.
* Each test file should test one function or a small group of related functions.
* Useful testing functions:
  * `expect_equal()`, `expect_identical()`, `expect_true()`, `expect_false()`
  * `expect_error()`, `expect_warning()`, `expect_message()`
* `test()` runs all tests.

## What to test

* Focus on testing the exported functions.
* Strive to test each behaviour in one and only one test.
* Avoid testing simple code that you’re confident will work.
* Always write a test when you discover a bug.
* The test-first philosophy: always start by writing the tests, and then write the code that makes them pass.
* Use `devtools::test_coverage()` to see which parts of your package are tested.

# Website

## pkgdown websites

* `usethis::use_pkgdown()`
  * Creates `_pkgdown.yml` to configure site.
  * Updates `.Rbuildignore`
  * Adds `docs` to `.gitignore`
* `pkgdown::build_site()` to build the site.
* `usethis::use_pkgdown_github_pages()` to publish the site via Gituhb Actiosn and GitHub Pages.
* Make a hex sticker with the `hexSticker` package.
* Add it using `usethis::use_logo()`.

# Continuous integration

## Github Actions
\vspace*{-0.4cm}

* Some development tasks can be executed automatically on Github with a trigger (e.g., a push)
* Run `R CMD check: `usethis::use_github_action("check_standard")`
* Compute test coverage: `usethis::use_github_action("test-coverage")`
* Build and deploy pkgdown site: `usethis::use_github_action("pkgdown")`
* The `.github/workflows/` directory contains action \rlap{files.}
* See [`https://github.com/r-lib/actions/`](https://github.com/r-lib/actions/) for more examples.
