---
title: "Week 5: Debugging"
---

```{r}
week <- 5
source(here::here("course_info.R"))
```


## Main reference

* [22 *Advanced R*](https://adv-r.hadley.nz/debugging.html)

## What you will learn this week

* Debugging

## Online resources

* [Creating a reprex](https://www.tidyverse.org/help/#reprex)
* [Posit Community](https://community.rstudio.com)
* [Stack Overflow](https://stackoverflow.com/)

## Pre-class activities

* Watch [Object of type 'closure' is not subsettable](https://www.youtube.com/watch?v=vgYS-F8opgE) with Jenny Bryan
* Watch [Debugging in R](https://www.youtube.com/watch?v=3YX-KyZ5Osw) with Jon Harmon



```{r}
#| output: asis
show_slides(week)
```

* [Survey data](survey_data.csv)
* [Exercise: MRE](survey_dplyr_bug.R)
* [Exercise: traceback](01-traceback.R)
* [Exercise: browser](02-browser.R)
* [Exercise: warnings](03-warnings.R)
* [Exercise: errors](04-errors.R)
* [Exercise: debug](05-debug.R)
* [Exercise: debug](06-debug.R)


```{r}
#| output: asis
show_assignments(week)
```
